Kaggle-in-class Data Challenges Can Boost Student Learning

### 2020-08-06 paper-kaggle.Rmd
comparison of UG and PG and tightening language in section 2

Computational Statistics and Data Mining (CSDM) is designed for postgraduate level  students with math, statistics, information technology or actuarial backgrounds. It covers modelling both continuous (regression) and categorical (classification) response variables. The 63 students were randomized into one of two kaggle competitions, one focused on regression (R) and the other classification (C). This setup mimics randomized control trials, which is the gold standard, in experiment design (@Shelley09CH1, ch. 1).
Students built prediction models and made submissions individually for 16 days, and then were allowed to form groups to compete for another 7 days.

@@

One of the 63 students elected not to take part in the competition, and another student didn't sit the exam, producing a final sample size of 61.

Table 1 shows some summary statistics for the two groups. We examine the percentage correct overall on the final exam for the different groups and the scores the students received for the second assignment. The second assignment examined students knowledge about computational methods and absolutely unrelated to the classification and regression methods. The difference between the two group are inconsiderable taking into the account the relatively small sample size of the groups.

@@

\begin{table}[h]
\begin{center}
\begin{tabular}{l l r r r r r r r}\hline
Score & Competition group & N & Mean (SD)   & Min  &  Q1 & Median  &  Q3  & Max \\\hline
Exam   & Regression      & 30 &  66.9 (14.1)  &   36 &  57.5 &   69.2 &  77.5 &   88.5\\
       & Classification & 31 &  62.1 (17.2)  &   20  & 52.2  &  68.0   &  75.2 &  85.5\\\hline

2$^{nd}$ assignment & Regression      &    30  & 8.16 (2.15)  &  0  & 7.02 &  8.35   &  9.9  &   10 \\
                    & Classification  &    31  & 8.01 (2.96)  &  0  & 7.8  &  9.30   &  9.9  &   10 \\\hline

\end{tabular}
\caption{Computational statistics and data mining: Basics summary statistics of the exam score (out of 100) and the second assignment (out of 10) for the two competition groups.}
\end{center}
\label{tab:Melb_Group_Compare}
\end{table}

@@

Similarly, to Table 1, Table 2 shows some summary statistics of the exam scores for the 34 postgraduate (ST-PG) students and for the 145 undergraduate (ST-UG) students. The mean and the median scores of postgraduate students are a bit lower than the corresponding scores of undergraduate students. However, the interquartile range is quite the same. Therefore, we do not have a principal evidence that postgraduate students are a stronger cohort. Interesting the highest exam score was received by an undergraduate student. Also, the mean and the median of the undergraduate students are a bit higher.

@@

\begin{table}[h]
\begin{center}
\begin{tabular}{l l r r r r r r r}\hline
Score & Competition group & N & Mean (SD)   & Min  &  Q1 & Median  &  Q3  & Max \\\hline
Exam   & ST-PG      & 34  & 61.8 (12.2)  &  35 & 57.2  &   61  & 72.2  & 81.5 \\
       & ST-UG         & 145 & 63.0 (15.4)  &   0 & 56.0  &   64  & 73.0   & 90.0\\\hline
\end{tabular}
\caption{Statistical thinking: Basics summary statistics of the exam score (out of 100) for the two groups.}
\end{center}
\label{tab:Melb_Group_Compare}
\end{table}

Figure \ref{fig:Monash_quiz} presents the means and the standard deviations of the quizzes taken by the undergraduate (ST-UG) and postgraduate (ST-PG) students.

Postgraduate students have slightly higher average marks on quizzes and much smaller standard deviation. However, the median is only higher by up to 1 point for postgraduate students (except for quiz 5, the difference is 2 points). It could suggest that they are a bit stronger cohort than the undergraduate students. But it also could be that they have a more serious attitude to the quizzes. Considering their performance in the exam the latter is more likely.

```{r Monash_quiz, fig.cap = "\\label{fig:Monash_quiz} Means and error bars of undergraduate (ST-UG) and postgraduate (ST-PG) students quizes during the semester.", fig.height=4.5, fig.width=6}
library(tidyverse)
library(gridExtra)
quiz <- read_csv(file="data/quiz_scores_by_unit_2420_5442.csv")

quiz <- quiz %>% mutate(unit = as.character(unit)) %>%
  mutate(quiz = factor(quiz, levels = c('Quiz 1','Quiz 2','Quiz 3','Quiz 4','Quiz 5','Quiz 6','Quiz 7','Quiz 8','Quiz 9','Quiz 10','Quiz 11','Quiz Tot')))

quiz1 <- quiz %>%   mutate(unit=replace(unit, unit=='2420','ST-UG')) %>%
  mutate(unit=replace(unit, unit=='5242','ST-PG'))

quiz1 %>% ggplot(aes(x=unit, y=mean, colour=unit)) +
  geom_point(aes(color=unit), size = 2) +
  geom_errorbar(aes(ymin=mean-sd, ymax=mean+sd), width=.2, 
                position=position_dodge(.9)) +
  facet_wrap(~quiz) + 
  labs(y="Score",x="") +
   theme(legend.position = "none")

@@

```{r Monash_quiz_mean_diff, fig.cap = "\\label{fig:Monash_quiz_mean_diff} Mean difference of undergraduate (ST-UG) and postgraduate (ST-PG) students quizes scores and CI.", fig.height=4.5, fig.width=6}

# Calculate CI for the differences in mean
t <- quiz %>% select(unit,quiz,mean,sd) %>%
      gather(statistics, value, -(quiz:unit)) %>%
      unite(temp, unit, statistics) %>%
      spread(temp, value)
t <- t %>% mutate(mean_dif = `5242_mean` - `2420_mean`,
                  sd = sqrt(`5242_sd`^2/34 + `2420_sd`^2/145),
                  v1 = `5242_sd`^2/34,
                  v2 = `2420_sd`^2/145,
                  v = (v1+v2)^2/(v1^2/(34-1)  + v2^2/(145-1) ),
                  t_q = qt( .975, df=v),
                  lower.CI = mean_dif - t_q*sd,
                  upper.CI = mean_dif + t_q*sd
)   
t %>% ggplot(aes(x=factor(quiz, levels = c('Quiz 1','Quiz 2','Quiz 3','Quiz 4','Quiz 5','Quiz 6','Quiz 7','Quiz 8','Quiz 9','Quiz 10','Quiz 11','Quiz Tot')) , 
                 y=mean_dif)) +
  geom_point( size = 2) +
  geom_errorbar(aes(ymin=lower.CI, ymax=upper.CI), width=.2, 
                position=position_dodge(.9)) +
  labs(y="Score",x="") +
   theme(legend.position = "none") +
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1))
```

```{r Monash_quiz_CI, fig.cap = "\\label{fig:Monash_quiz_CI} Means and CI bars of undergraduate (ETC2420) and postgraduate (ETC5242) students quizes during the semester.", fig.height=4.5, fig.width=6}

>>>>

Computational Statistics and Data Mining (CSDM) is designed for postgraduate level  students with math, statistics, information technology or actuarial backgrounds. It covers modelling both continuous (regression) and categorical (classification) response variables. The 63 students were randomized into one of two kaggle competitions, one focused on regression (R) and the other classification (C). (One of the 63 students elected not to take part in the competition, and another student didn't sit the exam, producing a final sample size of 61.) This setup mimics randomized control trials, which is the gold standard, in experiment design (@Shelley09CH1, ch. 1).
Students built prediction models and made submissions individually for 16 days, and then were allowed to form groups to compete for another 7 days.

@@

Table 1 compares the summary statistics for the two groups. We examine the percentage correct overall on the final exam for the different groups and the scores the students received for the second assignment. The second assignment examined students knowledge about computational methods, unrelated to the classification and regression methods. The two groups statistics are similar.

@@

\begin{table}[h]
\begin{center}
\begin{tabular}{l l | r | r | r r r r r}\hline
Score & Comp. group & N & Mean (SD)   & Min  &  Q1 & Median  &  Q3  & Max \\\hline
Exam   & Regression      & 30 &  66.9 (14.1)  &   36.0 &  57.5 &   69.2 &  77.5 &   88.5\\
       & Classification & 31 &  62.1 (17.2)  &   20.0  & 52.2  &  68.0   &  75.2 &  85.5\\\hline

2$^{nd}$ assignment & Regression      &    30  & 8.2 (2.2)  &  0.0  & 7.0 &  8.4   &  9.9  &   10.0 \\
                    & Classification  &    31  & 8.0 (3.0)  &  0.0  & 7.8  &  9.3   &  9.9  &   10.0 \\\hline

\end{tabular}
\caption{Computational statistics and data mining: Summary statistics of the exam score (out of 100) and the second assignment (out of 10) for the two competition groups.}
\end{center}
\label{tab:Melb_Group_Compare}
\end{table}

@@

Table 2 shows the summary statistics of the exam scores for the 34 postgraduate (ST-PG) students and for the 145 undergraduate (ST-UG) students. The mean and the median scores of postgraduate students are a bit lower than the corresponding scores of undergraduate students. However, the interquartile range is similar. Interestingly, the highest exam score was received by an undergraduate student.

@@

\begin{table}[h]
\begin{center}
\begin{tabular}{l l | r | r | r r r r r}\hline
Score & Competition group & N & Mean (SD)   & Min  &  Q1 & Median  &  Q3  & Max \\\hline
Exam   & ST-PG      & 34  & 61.8 (12.2)  &  35.0 & 57.2  &   61.0  & 72.2  & 81.5 \\
       & ST-UG         & 145 & 63.0 (15.4)  &   0.0 & 56.0  &   64.0  & 73.0   & 90.0\\\hline
\end{tabular}
\caption{Statistical thinking: Summary statistics of the exam score (out of 100) for the two groups.}
\end{center}
\label{tab:Melb_Group_Compare}
\end{table}

Figure \ref{fig:Monash_quiz} displays the quiz scores taken by the undergraduate (ST-UG) and postgraduate (ST-PG) students, as a beeswarm plot [@beeswarm]. The distribution of quiz scores are similar, except that undergaduates tend to have a longer lower tail.

<!-- Postgraduate students have slightly higher average marks on quizzes and much smaller standard deviation. However, the median is only higher by up to 1 point for postgraduate students (except for quiz 5, the difference is 2 points). It could suggest that they are a bit stronger cohort than the undergraduate students. But it also could be that they have a more serious attitude to the quizzes. Considering their performance in the exam the latter is more likely.-->

```{r Monash_quiz, fig.cap = "\\label{fig:Monash_quiz}Beeswarm plots of undergraduate (ST-UG) and postgraduate (ST-PG) students quizzes during the semester. The scores are very similar, although undergraduates have a longer lower tail.", fig.height=6, fig.width=6}
library(tidyverse)
library(gridExtra)
library(ggbeeswarm)
library(forcats)
quiz <- read_csv(file="data/quiz_scores_by_unit_2420_5242.csv")

quiz <- quiz %>% # mutate(unit = as.character(unit)) %>%
  mutate(quiz = factor(quiz, levels = c('Quiz 1','Quiz 2','Quiz 3','Quiz 4','Quiz 5','Quiz 6','Quiz 7','Quiz 8','Quiz 9','Quiz 10','Quiz 11','Quiz Tot'))) %>%
  mutate(unit = factor(unit, levels=c("ST_UG", "ST_PG"))) %>%
  group_by(quiz) %>%
  mutate(score = score/max(score)*100)

ggplot(quiz, aes(x=unit, y=score, colour=unit)) +
  geom_beeswarm(alpha=0.5) +
  facet_wrap(~quiz) + 
  xlab("") + ylab("Score (%)") +
  theme(legend.position="none")

#quiz1 %>% ggplot(aes(x=unit, y=mean, colour=unit)) +
#  geom_point(aes(color=unit), size = 2) +
#  geom_errorbar(aes(ymin=mean-sd, ymax=mean+sd), width=.2, 
#                position=position_dodge(.9)) +
#  facet_wrap(~quiz) + 
#  labs(y="Score",x="") +
#   theme(legend.position = "none")

@@

```{r Monash_quiz_mean_diff, fig.cap = "\\label{fig:Monash_quiz_mean_diff} Mean difference of undergraduate (ST-UG) and postgraduate (ST-PG) students quizes scores and CI.", fig.height=4.5, fig.width=6, eval=FALSE}

# Calculate CI for the differences in mean
t <- quiz %>% select(unit,quiz,mean,sd) %>%
      gather(statistics, value, -(quiz:unit)) %>%
      unite(temp, unit, statistics) %>%
      spread(temp, value)
t <- t %>% mutate(mean_dif = `5242_mean` - `2420_mean`,
                  sd = sqrt(`5242_sd`^2/34 + `2420_sd`^2/145),
                  v1 = `5242_sd`^2/34,
                  v2 = `2420_sd`^2/145,
                  v = (v1+v2)^2/(v1^2/(34-1)  + v2^2/(145-1) ),
                  t_q = qt( .975, df=v),
                  lower.CI = mean_dif - t_q*sd,
                  upper.CI = mean_dif + t_q*sd
)   
t %>% ggplot(aes(x=factor(quiz, levels = c('Quiz 1','Quiz 2','Quiz 3','Quiz 4','Quiz 5','Quiz 6','Quiz 7','Quiz 8','Quiz 9','Quiz 10','Quiz 11','Quiz Tot')) , 
                 y=mean_dif)) +
  geom_point( size = 2) +
  geom_errorbar(aes(ymin=lower.CI, ymax=upper.CI), width=.2, 
                position=position_dodge(.9)) +
  labs(y="Score",x="") +
   theme(legend.position = "none") +
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1))

```{r Monash_quiz_CI, fig.cap = "\\label{fig:Monash_quiz_CI} Means and CI bars of undergraduate (ETC2420) and postgraduate (ETC5242) students quizes during the semester.", fig.height=4.5, fig.width=6, eval=FALSE}
